---
title: "Class 22 - In class exercises"
author: Hari Subhash
date: "Date created: `r Sys.Date()`"
output:
  html_notebook:
    highlight: haddock
    smart: yes
    theme: readable
    df_print: paged
---
## Functions
The code below creates a function that takes two numeric vectors as its input and returns the absolute difference between the minimum value of the first vector and the maximum value of second vector. Read it carefully (along with the comments).

```{r}
##Defining the function (notice how I also define the na.rm.)
minmax_diff <- function(min_vec, max_vec, na.rm = F) {
  ##check if the two arguments are of the type numeric, else exit.
  stopifnot(is.numeric(min_vec), is.numeric(max_vec), is.logical(na.rm))
  
  ##get the min and max value.
  min_value <- min(min_vec, na.rm = na.rm)
  max_value <- max(max_vec, na.rm = na.rm)
  
  ##return the absolute difference
  abs(max_value - min_value)
  
}
```

Let's test to see if it works.
```{r}
minmax_diff(min_vec = 1:10, max_vec = 10:20)
```

Introducing NA values.
```{r}
minmax_diff(min_vec = c(24, 41, NA), max_vec = 44)
```

Specifiying the NA condition.
```{r}
minmax_diff(min_vec = c(24, 41, NA), max_vec = 44, na.rm = T)
```

Trying out a non-numeric input.

```{r}
minmax_diff(min_vec = "a", max_vec = c(1, 2, 56))
```

1. How does this code work?
```{r}
minmax_diff(1:23, 4, T)
```

2. Why does this give an error? 

```{r}
minmax_diff(1, 2, 5)
```

3. Why does this not work?

```{r}
minmax_diff(1, 2, F, 6)
```

4. Define a function that takes two numeric vectors as its input and returns the difference between the mean of the first vector and the median of the second vector. Add any checks and additional arguments you like to this function.
5. Define a function that takes two arguments - name and greet_message. Set the default value of greet_message to be "Hello". The function should return a single character string that combines greet_message with name (with a space in between). Use `paste()` to combine. Add any additional arguments or checks.
6. Load the tidyverse and gapminder packages
7. Fix the following so that they run. Think about why they are generating an error. Read the comments for further instructions.
```{r}
##Find the unique set of continents in gapminder
##Your final solution should use pipes
gapminder %>% 
  unique(continent)
```
```{r}
##find the mean of the gdpPercap column in gapminder
##final solution should use pipes.
gapminder %>% 
  select(gdpPercap) %>% 
  mean(., na.rm = T)
```


```{r}
## find the mean of the gdpPercap column in gapminder
## final solution should use base R syntax 
## remember [] - is a preserving subset | $ and [[]] are simplifying
mean(gapminder["gdpPercap"])
```

```{r}
## I would like the distinct combinations of continent and country
## use tidyverse
gapminder %>% 
  unique(continent, country)
```

```{r}
## I would like to have a single column tibble that contains all the distinct continents
## Use tidyverse
gapminder %>% 
  pull(continent) %>% 
  distinct()
```

