---
title: "ss_v_midclass"
author: "Mario"
date: 'Date created: `r Sys.Date()`'
output:
  html_notebook:
    code_folding: hide
    df_print: paged
    highlight: kate
    smart: yes
    theme: paper
    toc: no
  html_document:
    toc: no
---
# {.tabset} 

## abstract

ojfowj

## background

Empirically, Social Security is one of the U.S.'s most successful social programs.

## setup

Install and load packages
```{r message=FALSE, message=FALSE, warning=FALSE}

# install.packages('gapminder'); install.packages('DiagrammeR')
library(tidyverse); library(lubridate); library(gapminder); library(DiagrammeR); library(ggthemes)
```

Load poverty guideline data
```{r}
poverty_hist <- read.csv(file = 'hstpov1.csv', header = T, na.strings = '***') %>%
  mutate(yr = year
         ,yr_mth = yr + 0.04) %>%
  select(yr_mth, X2_senior, X1_senior, X4_all)
```

Load social security cost of living adjustment data
```{r}
ss_cola <- read.delim(sep = ' '
             ,file = 'cola_1975-2019.txt'
             ,strip.white = T
             ,header = F
             ,col.names = c(1:4,'mth_name', 'yr', 'rmv', 'cola')) %>%
  mutate(mth_name = trimws(mth_name)
         ,mth = match(mth_name, month.name)
         ,yr_mth = yr + round(((mth - 0.5) / 12), 2)
         ,cola = as.numeric(sub(x= cola, '%', '')) / 100) %>%
  select(yr_mth, cola)


```

Load social security historic benefits data
```{r}

socsec_hist <- read.csv(file = 'some_socsec.csv', header = T) %>%
  mutate(month_day_year = mdy(month) 
         ,yr = year(month_day_year)
         ,mth = month(month_day_year)
         ,yr_mth = yr + round(((mth - 0.5) / 12), 2)) %>%
  select(yr_mth, mean_ss)
```


Load average household size historic data
```{r}

house_hist <- read.csv(file = "hh4.csv", skip = 2, header = T) %>%
  mutate(yr_mth = as.numeric(str_sub(year,1,4)) + 0.04) %>%
  select(yr_mth, mean_house_size) %>%
  arrange(yr_mth)
```


Load median income data. This is the final load.
```{r}

mdn_y_hist <- read.csv(file = 'mdn_house-fam_y.csv', na.strings = 'NULL') %>%
  mutate(yr = if_else(year(mdy(DATE)) < 2020
                      ,year(mdy(DATE))
                      ,year(mdy(DATE)) - 100)
         ,mth = month(mdy(DATE))
         ,yr_mth = yr + round(((mth - 0.5) / 12), 2)) %>%
  select(yr_mth, mdn_house_y, mdn_fam_y)
```

### Standardizing the data

To test the hypothesis, I'll standardize the numbers for two elderly recipients for a whole year. There are poverty guidelines for each size of household and elderly-specific guidelines for a household of two, so those can be loaded as is.
```{r}

my_work <- poverty_hist %>%
  mutate(poverty_2seniors = X2_senior) %>%
  select(yr_mth, poverty_2seniors)
```

Average Social Security benefits received is for only one recipient and is monthly, so will need to be doubled and multiplied for 12 months.
```{r}

my_work <- my_work %>%
  full_join(y = socsec_hist %>%
              mutate(mean_ss_2seniors = 2 * 12 * mean_ss) %>%
              select(yr_mth, mean_ss_2seniors)
            ,by = 'yr_mth')
```

Median household income is given as an annual amount, but the size of a household varies by year. I'll divide by mean household size then multiply by 2
```{r}

my_work <- mdn_y_hist %>%
  full_join(house_hist, by = 'yr_mth') %>%
  mutate(mdn_y_2seniors = mdn_house_y * 2 / mean_house_size) %>%
  full_join(my_work, by = 'yr_mth') %>%
  arrange(yr_mth) %>%
  select(yr_mth, mdn_y_2seniors, poverty_2seniors, mean_ss_2seniors) %>%
  filter(!is.na(mdn_y_2seniors) | !is.na(poverty_2seniors | !is.na(mean_ss_2seniors)))
```

Now to see whether seniors have ever been middle class, and if so, whetgher remained middle class
```{r}

my_work %>%
  mutate(ss_as_mdn = mean_ss_2seniors / mdn_y_2seniors
         ,pov_as_ss = poverty_2seniors / mean_ss_2seniors) %>%
  ggplot(mapping = aes(yr_mth, mdn_y_2seniors)) +
  geom_point() +
  geom_line(mapping = aes(yr_mth, mean_ss_2seniors), color = 'darkred') +
  theme(legend.position="right")

##  theme_fivethirtyeight() +
##  legend(legend = 'ss_as_mdn', x = 'right')
```

