---
title: "Class 10 - In class exercises"
author: Hari Subhash
date: "Date created: `r Sys.Date()`"
output:
  html_notebook:
    highlight: kate
    smart: yes
    theme: cosmo
    code_folding: show
    df_print: paged
---

##Task 1: Examine the weather data
1. Use the `skim()` function from the `skimr` package to visually display all the variables in weather

```{r}
install.packages('skimr')
library(skimr);library(nycflights13);library(tidyverse)

skim(weather)
```


2. Use `nrow()` to display the number of rows in the weather data

```{r}
nrow(weather)
```


3. Use `dim()` to display the rows and columns of the weather data
```{r}
dim(weather)
```

4. Use `names()` to display the column names of the weather data
```{r}
names(weather)
```

5. Display the `head()` of the weather data

```{r}
head(weather)
```


6. Display the `tail()` of the weather data

```{r}
tail(weather)
```

7. Display a random 10 rows from the weather data using `sample_n()`

```{r}
sample_n(size = 10,weather)
```


8. What are the primary keys of the weather data? i.e. which column(s) can be used to uniquely identify each observation in the weather data? Hint: `group_by()` with `filter(n() > 1)` to check if there are more than one observation per group.

```{r}
weather %>%
  group_by(time_hour,origin) %>%
  filter(n()>1)

weather %>%
   arrange(desc(precip)) %>%
  select(1,3:5,precip)
```


##Task 2: Explore the indepedent variable
Use a combination of text, tables and charts to explore the nature of the independent variable

```{r}
summary(weather$wind_gust);str(weather$wind_gust); ggplot(data = weather, aes(wind_gust)) + geom_histogram(bins = 20); ggplot(data = weather, aes(wind_gust)) + geom_density(); weather %>% 
    mutate(binned_gust = cut_width(wind_gust, width = 5)) %>% 
    group_by(binned_gust) %>% 
    summarise(nObs = n())


weather %>% 
  filter(wind_gust > 0) %>%
  arrange(wind_gust) %>% 
  select(1:5,wind_gust) %>%
  group_by(wind_gust) %>% 
  summarize (nObs = n()) %>%
  mutate(wind_gust_kph = round(wind_gust * 4 * 1.60921449)/4) %>%
  arrange(wind_gust)



```



##Task 3. Explore a relationship
Explore the relationship between departure delay and a variable of your choice.

```{r}
str(weather$visib);summary(weather$visib);skim(weather$visib)

weather %>%
  filter(visib<10) %>%
  ggplot(aes(visib)) +
  geom_histogram(bins = 50)



weather %>%
  group_by(visib) %>%
  summarise(mean_)
  ggplot(aes(wind_gust,visib)) +
  geom_point()

delayed_flights<- flights %>%
  filter(dep_delay>0) %>%
  select(time_hour,origin,dep_delay)

weather %>%
  select(time_hour,origin,visib) %>%
  inner_join(delayed_flights, by = c('time_hour','origin')) %>%
  group_by(visib) %>%
  summarize(mean_delay = mean(dep_delay,na.rm = T)) %>%
  ggplot(aes(visib,mean_delay)) +
  geom_point() +
  geom_smooth(method = loess)
```

##Bonus task! Explore a relatioship between two independent variables

```{r}
library(nycflights13);library(tidyverse);library(skimr)

weather %>%
  select(visib,precip) %>%
  group_by(visib) %>%
  summarize(mean_precip = mean(precip)) %>%
  ggplot(aes(visib,mean_precip)) +
  geom_point() +
  geom_smooth()

weather %>%
  select(visib,precip) %>%
  group_by(visib) %>%
  summarize(mean_precip = mean(precip)) %>%
  ggplot(aes(visib,mean_precip)) +
  geom_point() +
  geom_smooth()

weather %>%
  select(time_hour,origin,visib,precip) %>%
  inner_join(delayed_flights, by = c('time_hour','origin')) %>%
  group_by(visib) %>%
  mutate(mean_delay = mean(dep_delay,na.rm = T)) %>%
  ggplot(aes(visib,mean_delay,colors=precip)) +
  geom_point() +
  geom_smooth(method = loess)



```

