---
title: "Assignment 2"
author: "~~Hari Subhash~~ Mario"
date: 'Date created: `r Sys.Date()`'
output:
  html_notebook:
    code_folding: show
    df_print: paged
    highlight: kate
    smart: yes
    theme: paper
    toc: no
  html_document:
    toc: no
---

# {.tabset .tabset-fade .tabset-pills}

## Abstract 

Unpredictable flight delays are a costly problem, forcing travelers to choose between spending unproductive time at the airport and missing a flight. In 2013, flights from the three New York area airports alone saw aggregate delays of half a million minutes. I explore the predictive capacity with respect to departure delay of five variables -- **scheduled departure time**, **humidity**, **atmomspheric pressure**, **the probability associated with each airplane model**, and **the probability associated with each carrier**. Notably all five of these variables are forecasted ahead of time and made publicly available. Of those five variables, it's **scheduled departure time** that is most closely associated with delay.

## Background

**PROBLEM:** Here, the problem is that flight delays are unpredictable. Not only that, mis-timing your ride to the airport could be costly; Arrive late and you may miss your flight altogether, but arrive early and you'll spend even more uncomfortable, unproductive time at the gate. In an ideal world, all flights would simply leave on time. In a second-best world, each airline would contact travelers directly with flight delay information that was accurate and timely enough to be actionable. Alas, neither of those are the case.

**GOAL:** My ultimate goal is to be able to best predict whether a flight will be delayed using weather, plane, and flight data.

**HYPOTHESIS:** My null hypothesis is that *no* available weather, plane, or flight data shares a relationship with flight delay. I'll seek to disprove this.

## Setting up

To set up, I'll do the following:

1. Link flight data with data to the planes flown and the weather contemporaneous to each flight
2. Clean and adapt the data both so that it makes sense logically and is testable; This will include:
    a. Setting negative departure delay values equal to zero^1^
    b. Replacing categorical data -- such as the name of an airline -- with an "object effect" that estimates the added chance for delay associated with any object
3. Prepare an appendix of metadata, including:
    a. A data dictionary
    b. A data lineage diagram


```{r}
### On the code side, I'll do the following:

# 1. Install and enable R packages to explore, manipulate, and analyze the data
# 2. Use common fields to join flight table w/weather and plane tables
# 3. Select columns to be removed
# 4. Mutate some columns so they're more useful. Specifically, I group categorical variables and estimate a conditional probability using observed delay data

### 1. Install packages, as needed 
# install.packages('tidyverse'); install.packages('skimr'); install.package('lubridate');
# install.packages('ggthemes'); install.packages('DiagrammeR')

# This line enable packages
library(tidyverse); library(skimr); library(nycflights13); library(lubridate); library(ggthemes)
```
```{r}
### 2. Join flight, plane, and weather data into a single data frame of raw data to work with
flt_raw <- flights %>%
  left_join(rename(planes,'manuf_year' = year), by = 'tailnum') %>% ## Rename plane manufacture year from 'year' to 'manuf_year' to not confuse it with flight year
  left_join(weather, by = c('time_hour', 'origin'))
```


```{r}
### Remove irrelevant columns and transform categorical data into numbers


flt_test <- flt_raw %>%
  filter(!is.na(dep_delay) == T) %>%
  select(dep_delay, sched_dep_time, distance, manuf_year, engines ## 1-5
         ,seats, temp, dewp, humid, wind_speed,wind_gust, precip ## 6-12
         ,pressure, visib, distance ## 13-15
         ,carrier, origin, type, manufacturer ## cat cols 16-19
         ,model, speed ,engine) %>% ## cat cols 20-22
  mutate(diff_20h = if_else(sched_dep_time < 2000
                            ,1960 - sched_dep_time
                            ,sched_dep_time - 2000)
                            ,dep_delay = if_else(dep_delay > 0, 1, 0)) %>%
  group_by(carrier) %>%
     mutate(op_carr = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup() %>%
  group_by(origin) %>%
     mutate(op_ogn = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup() %>%
  group_by(type) %>%
     mutate(op_type = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup() %>%
  group_by(manufacturer) %>%
     mutate(op_mnf = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup() %>%
  group_by(model) %>%
     mutate(op_mdl = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup() %>%
  group_by(speed) %>%
     mutate(op_spd = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup() %>%
  group_by(engine) %>%
     mutate(op_eng = sum(dep_delay>0) / length(dep_delay) - op_base) %>%
     ungroup()
```

```{r}
### This code creates a separate data frame which links categorical variables to their relative probabilities
cat_facts <- flt_test %>%
  select(15:21,23:29) %>%
  unique()

## Now, I'll gather and do a little clean-up

cat_facts <- cat_facts %>%
  gather(key = 'cat', value = 'obj', carrier, origin, type, manufacturer, model, speed, engine) %>%
  rename(carrier = op_carr
         ,origin = op_ogn
           ,type = op_type
   ,manufacturer = op_mnf
          ,model = op_mdl
          ,speed = op_spd
         ,engine = op_eng) %>%
  gather(key = 'trans', value = 'facts', 1:7) %>%
  filter(cat == trans) %>%
  select(-trans) %>%
  unique() %>%
  filter(!is.na(obj) == T) %>%
  arrange(cat,desc(facts))

cat_facts
```  
  
```{r}
### This code removes the original categorical variables to get the data frame even skinnier; Can reference cat_facts if needed

flt_test <- flt_test %>%
  select(-(15:21))

flt_test %>%
  cor(use = 'pairwise.complete.obs') %>%
  as_tibble() %>%
  mutate(field = rownames(cor(flt_test))) %>%
  filter(dep_delay < 1) %>%
  select(field, dep_delay) %>%
  arrange(desc(abs(dep_delay)))

glm(op_mdl, na.action = na.exclude(), family = binomial, data = flt_test)

```

^1^ A negative departure delay value would decrease a category mean as though it *is* more on time.  A flight that leaves early should not be considered "more on time" than a flight that leaves at it's scheduled time. To better intuit this, suppose an airline has one flight that leaves 30 minutes early and another that leaves 30 minutes late. On paper it would appear that a traveler need not worry about delays when flying on this airline, but this is not the case. Because of that, I chose to zero out negative departure delay values.

## Univariate analysis

Here are the steps for this

1. Explore metadata to ensure I understand what each variable represents
2. Descriptive statics of the dependent variable, departure delay, including number of observations, mean, standard deviation, median, min, max, and interquartile measures
3. Descriptive statistics for the independent variables
4. A histogram will offer a visual representation of each

Doing this will expose:

* Fields that would be unlikely related to departure delay,
* Fields that will require transformation, and
* Fields with too few observations or with too many missing observations to be representative

**DATA DICTIONARY**

* **year,month,day:** Date of departure or date of weather recording
* **dep_time,arr_time:** Actual departure and arrival times (format HHMM or HMM), local tz.
* **sched_dep_time,sched_arr_time:** Scheduled departure and arrival times (format HHMM or HMM), local tz.
* **dep_delay,arr_delay:** Departure and arrival delays, in minutes. Negative times represent early departures/arrivals.
* **hour** Hour part of time of scheduled departure or weather recording
* **minute:** Minute part of time of scheduled departure
* **carrier:** Two letter carrier abbreviation. See airlines() to get name
* **tailnum:** Plane tail number
* **flight:** Flight number
* **origin:** Origin of flight or weather station. Origin + time_hour can be used to join flight and weather data
* **dest:** Destination of flight. See airports() for additional metadata.
* **air_time:** Amount of time spent in the air, in minutes
* **distance:** Distance between airports, in miles
* **time_hour:** Scheduled date and hour of the flight or of the weather recording as a POSIXct date. Origin + time_hour can be used to join flight and weather data
* **temp,dewp:** Temperature and dewpoint in F
* **humid:** Relative humidity
* **wind_dir,wind_speed,wind_gust:** Wind direction (in degrees), speed and gust speed (in mph)
* **precip:** Precipitation, in inches
* **pressure:** Sea level pressure in millibars
* **visib:** Visibility in miles

**DESCRIPTIVE STATS OF THE DEPENDENT VARIABLE:** 

First, I'll prepare a histogram of departure delay. I'll cut off at the 97.5th percentile so it's easier to see

```{r}
flights %>% ##
  select(dep_delay) %>%
  filter(!is.na(dep_delay) == T
           & dep_delay < as.numeric(quantile(flights %>%
           filter(!is.na(dep_delay) ==T) %>%
           select(dep_delay) %>%
           as_vector(), probs = .975))) %>%
  mutate(dep_delay = (dep_delay + abs(dep_delay)) / 2) %>%
  as_tibble() %>%
  ggplot(data = ., mapping = aes(x = dep_delay)) +
  geom_histogram(binwidth = 5, na.rm = T) +
  theme_economist() +
  ggtitle('Departure delay is skewed right and not normally distributed')
  
```
Let's look at some percentiles to verify that this is not normal by checking the probabilities associated with one-tailed z-scores of 0.5 to 3.0
```{r}
quantile(flights %>%
           filter(!is.na(dep_delay) ==T) %>%
           select(dep_delay) %>%
           mutate(dep_delay = (dep_delay + abs(dep_delay))/2) %>%
           as_vector(), probs = c(0.6915, 0.8413,0.9332 ## p(z < 0.5, 1.0, 1.5)
                                  ,0.9772, 0.9938, 0.9987)) ## p(z < 2.0, 2.5, 3.0)
```
In a normal distribution, the distance between each of these would be equal. 

```{r}




```


```{r}
### Here, I'll calculate the baseline probability of delay
op_base <- flt_raw %>%
  filter(!is.na(dep_delay) == T) %>%
  group_by(dep_delay > 0) %>%
  summarise(p = length(dep_delay)/ sum(!is.na(flights$dep_delay) == T) ) %>%
  filter(p<.5) %>%
  select(2) %>%
  as.numeric()

# I'll use this to normalize individual category probabilities

```
The following fields are either unlikely to be related to departure delay or duplicative of another, superior field in the data set

* Destination
* Tail number
* Arrival time
* Day of the month
* Flight number
* Scheduled arrival time
* Year
* Air time
* Arrival delay
* Departure time
* Distance

Only one field among those remaining, **wind gust**, is concerning due to the number of n/a entries -- 256,391 of 336,776 total. My fear is that these incomplete entries are biased -- that is, we may be telling a story that is based on a sample of the data that does not represent the poopulation.
```{r}
### Here, I split up all entries based on whether or not wind_gust is NULL
### Then, look at descriptive statistics of delay by those groups
flt %>%
  select(wind_gust,dep_delay) %>%
  filter(is.na(dep_delay) == F) %>%
  mutate(na_gust = if_else(is.na(wind_gust), 0, 1)) %>% 
  group_by(na_gust) %>% 
  summarise(p_del = sum(dep_delay > 0)/length(dep_delay) 
     ,n_del     = length(dep_delay)
     ,min_del    = min(dep_delay,na.rm = T)
     ,del_25qile = quantile(dep_delay,probs = 0.25,na.rm = T)
     ,mdn_del    = median(dep_delay,na.rm = T)
     ,del_75qile = quantile(dep_delay,probs = 0.75,na.rm = T)
     ,max_del    = max(dep_delay, na.rm = T)) %>%
    round(2)

flt %>%
  select(wind_gust,dep_delay) %>%
  filter(is.na(dep_delay) == F) %>%
  mutate(na_gust = factor(if_else(is.na(wind_gust), 0, 1)
                             ,levels = c(0,1)
                             ,labels = c('null', 'not null') )) %>%
  ggplot(aes(dep_delay)) +
  geom_density() +
  ggtitle(label = 'Dep. delay differs when wind gust is na but is good enough to use') +
  theme_economist() +
  facet_wrap(~na_gust, nrow = 2, ncol = 1)
```
These look similar enough to work with. To verify with numbers, I would do a z-test to see if I'm able to disprove that the means are unequal and an f-test to do the same with the standard deviations. But not today. Today, I will do neither.

Now to filter out the unused fields and transform other fields:
```{r}
library(nycflights13)

flt_wx_cln <- flt_wx_raw %>%
  group_by(carrier) %>%
  mutate(p_carr = sum(dep_delay > 0) / sum(!is.na(dep_delay)) - 0.39094 ## Creating the carrier effect
         ,is_wkend   = if_else(wday(as.Date(time_hour)) == 1
            | wday(as.Date(time_hour)) == 7,1,0) ## Creating categorical variable for weekend-ness
         ,diff_20hr  = if_else(sched_dep_time < 2000, 1960 - sched_dep_time, sched_dep_time - 2000) ## Transforming time variable by difference from 8 p.m.
         ,diff_7mo   = abs(month.x-7)) %>%  ## Transforming month variable by difference from July
  ungroup() %>%
  group_by(origin) %>%
  mutate(p_ogn = sum(dep_delay > 0) / sum(!is.na(dep_delay)) - 0.39094) %>% ## Deriving origin effect
  ungroup() %>%
  select(dep_delay, temp:p_ogn)



skim(flt_wx_cln)
```
## Bivariate analysis

There are steps here

1. Scatter plots to intuit linear relationships to explore
2. Find correlation
3. Linear model the best

**FINDING LINEAR RELATIONSHIPS:** I'm looking for scatter plots that appear generally linear with as little spread around the line as possible

```{r, message=FALSE, warning=FALSE}
flt_wx_cln %>%
  group_by(temp) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = temp, y = mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Temp appears to have a relationship albeit weak')

flt_wx_cln %>%
  group_by(dewp) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = dewp, y = mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Dewpoint looks real similar to temp which makes sense')

flt_wx_cln %>%
  group_by(humid) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = humid, y = mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Humidity is the best so far but still weak')

flt_wx_cln %>%
  group_by(wind_speed) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = wind_speed, mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Std dev increases as wind speed increases BAD')

flt_wx_cln %>%
  group_by(wind_gust) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = wind_gust, mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Wind gust is weak')

flt_wx_cln %>%
  group_by(precip) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = precip, mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Precipitation looks OK')

flt_wx_cln %>%
  group_by(pressure) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = pressure, mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Pressure looks pretty decent')

flt_wx_cln %>%
  group_by(visib) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = visib, mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Visibility is on the good side of OK')

flt_wx_cln %>%
ggplot(aes(x = carr_delay, dep_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_economist() +
  geom_jitter() +
  ggtitle('Carrier effect graphs poorly but could be nice')

flt_wx_cln %>%
ggplot(aes(x = jitter(is_wkend, 0.8), jitter(dep_delay,10))) +
  geom_point() +
  geom_smooth(method = loess, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Weekend day also graphs poorly')

flt_wx_cln %>%
  group_by(diff_20hr) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = T)) %>%
ggplot(aes(x = diff_20hr, mean_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Time difference with 8 pm looks like the best')

flt_wx_cln %>%
ggplot(aes(x = diff_7mo, dep_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  ggtitle('Month difference with July isnt great')

flt_wx_cln %>%
ggplot(aes(x = ogn_delay, dep_delay)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  theme_economist() +
  ggtitle('Origin effect also is not good')




# ASCII bits for building in-line histograms
# 0| |▁
# 1|▂|▃
# 2|▄|▅
# 3|▆|▇
# 4|█|

```

The variables that appear to have the best relationships with dep_delay are:

1. diff_20hr (the best by far),
2. humid,
3. pressure,
4. carr_delay, and
5. dewp

One initial concern I have is that humidity and dewpoint are telling the same part of the story, but that's not so big a concern .

For now, let's shrink our working data frame for the last time to the remaining variables we still care about.
```{r}
flt_wx <- flt_wx_cln %>%
  select(dep_delay, diff_20hr, humid, pressure, carr_delay, dewp)

flt_wx
```


**OUTLIERS:** 

```{r message=FALSE, warning=FALSE}
flt_wx %>%
  mutate(cats = cut_number(diff_20hr, n = 5)) %>%
  ggplot(aes(group = cats, x = cats, y = dep_delay)) +
  geom_boxplot() +
  geom_jitter(width = 0.49, height = 0.49) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight()

flt_wx %>%
  mutate(cats = cut_number(humid, n = 5)) %>%
  ggplot(aes(group = cats, x = cats, y = dep_delay)) +
  geom_boxplot() +
  geom_jitter(width = 0.49, height = 0.49) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight()

flt_wx %>%
  mutate(cats = cut_number(pressure, n = 5)) %>%
  ggplot(aes(group = cats, x = cats, y = dep_delay)) +
  geom_boxplot() +
  geom_jitter(width = 0.49, height = 0.49) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight()

flt_wx %>%
  mutate(cats = cut_number(carr_delay, n = 5
                           ,labels = c('carr_delay_1', 'carr_delay_2', 'carr_delay_3', 'carr_delay_4', 'carr_delay_5'))) %>%
  ggplot(aes(group = cats, x = cats, y = dep_delay)) +
  geom_boxplot() +
  geom_jitter(width = 0.49, height = 0.49) +
  theme_igray()

flt_wx %>%
  filter(is.na(dewp) == F) %>%
  mutate(cats = cut_number(dewp, n = 5
                           ,labels = c('dewp_1', 'dewp_2', 'dewp_3', 'dewp_4', 'dewp_5'))) %>%
  ggplot(aes(group = cats, x = cats, y = dep_delay)) +
  geom_jitter(width = 0.49, height = 0.49) +
  scale_color_fivethirtyeight() +
  theme_fivethirtyeight() +
  geom_boxplot()
```


**CORRELATION:** Now that we've picked our independent variables, I'll verify using correlation

Correlation measures the degree to which two variables move up or down in relation to each other. Correlation must fall somewhere between -1.00 and 1.00, which imply a perfectly negative relationship and a perfectly positive relationship, respectively.


And now correlation.
```{r}
flt_wx %>%
  cor(use = 'pairwise.complete.obs', method = 'pearson') %>%
  round(2) %>%
  as.data.frame() %>%
  select(1) %>%
  cbind(rownames(.),.) %>%
  rename('variable' = 'rownames(.)'
         ,'corr_dep_delay' = 'dep_delay') %>%
  filter(corr_dep_delay > -1 & corr_dep_delay < 1) %>%
  arrange(desc(abs(corr_dep_delay)))
```
As predicted, diff_20hr has the best relationship with corr_dep_delay. Though it doesn't appear that the magnitude of any correlations is very high.

**LINEAR MODEL:** Having confirmed diff_20hr as the variable to use, we can let R build a linear model.

```{r}
lm(formula = dep_delay ~ dif, data = flt_test)
```




## Multivariate analysis

1. Covariance
2. Linear model multiple

